---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---

```{r setoptions, echo=FALSE}
## Set global options
##install.packages("knitr")
##install.packages("weatherData",repos = "http://cran.us.r-project.org")
##library(knitr) 
##opts_chunk$set(echo=TRUE, results="asis")
```

## Step 1: Code for reading in the dataset and/or processing the data

```{r, echo=TRUE}
## Load and preprocess the data 
acty_data1 <- read.csv2(unz("activity.zip", "activity.csv"), header=TRUE,sep=",")
acty_data1$steps <- as.numeric(acty_data1$steps) 
acty_data1$date <- as.Date(acty_data1$date) 
acty_data1 <- transform(acty_data1, date=factor(date), interval=factor(interval))
str(acty_data1)
head(acty_data1)
```

## Step 2: Histogram of the total number of steps taken each day

```{r}
## Summarize number of steps by date, excluding NA values
acty_data2 <- aggregate(x=acty_data1[1], by=acty_data1[c("date")], FUN=sum, na.rm=TRUE)
## Histogram of the total number of steps taken each day
hist(acty_data2$steps, xlab="Steps", main="Total number of steps taken each day")
```

## Step 3: Mean and median number of steps taken each day

```{r}
## Calculate mean and median number of steps taken each day
library(doBy)
acty_data3 <- summaryBy(steps ~ date, data = acty_data1, FUN = list(mean, median), na.rm=TRUE)
str(acty_data3)
head(acty_data3)
```

## Step 4: Time series plot of the average number of steps taken

```{r}
## Plot of the average number of steps taken by date, including NA
acty_data3 <- transform(acty_data3, date=factor(date))
library(ggplot2)
ggplot(acty_data3, aes(date, steps.mean, group = 1)) + geom_point() + geom_line() +
  labs(x = "Date", y = "Steps", title = "Average number of steps taken each day (w NA)")
```

## Step 5: The 5-minute interval that, on average, contains the maximum number of steps

```{r}
## Exclude NA rows
acty_data4 <- acty_data1[complete.cases(acty_data1),]
## Create numeric column for number of steps
max_step <- max(acty_data4$steps, na.rm=TRUE)
## Search for row with max steps
acty_data4[acty_data4$steps == max_step, ]
## Here is an alternative approach using arrange function; might be more efficient
library(dplyr)
head(arrange(acty_data1, desc(steps)),1)
```

## Step 6: Code to describe and show a strategy for imputing missing data

```{r}
## Calculate and report the total number of missing values in the dataset 
sum(is.na(acty_data1$steps))
## Imputing missing values.  For all the steps value as NA, set to 0
## This is no ideal solution
acty_data1$stepsImputed <- acty_data1$steps
acty_data1[is.na(acty_data1$steps),4] <- 0
str(acty_data1)
head(acty_data1)
sum(is.na(acty_data1$stepsImputed))
```

## Step 7: Histogram of the total number of steps taken each day after missing values are imputed

```{r}
## Plot of the average number of steps taken by date, excluding NA
acty_data3A <- summaryBy(stepsImputed ~ date, data = acty_data1, FUN = list(mean, median), na.rm=TRUE)
acty_data3A <- transform(acty_data3A, date=factor(date))
ggplot(acty_data3A, aes(date, stepsImputed.mean, group = 1)) + geom_point() + geom_line() + 
  labs(x = "Date", y = "Steps", title = "Average number of steps taken each day (Imputed)")
```

## Step 8: Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r}
## Calculate weekday_type column based on date
acty_data1$weekday <- weekdays(as.Date(acty_data1$date))
acty_data1$weekday_type <- sapply(acty_data1$weekday, switch,
                                  Monday ='Weekday', Tuesday ='Weekday', Wednesday ='Weekday', 
                                  Thursday ='Weekday', Friday = 'Weekday', Saturday = 'Weekend', 
                                  Sunday = 'Weekend')
acty_data1 <- transform(acty_data1, weekday_type=factor(weekday_type), interval=factor(interval))
## Summarize average number of steps by weekday_type and interval, excluding NA values
acty_data5 <- aggregate(x=acty_data1[1], by=acty_data1[c("weekday_type","interval")], FUN=mean, na.rm=TRUE)
str(acty_data5)
head(acty_data5)

library(ggplot2)
ggplot(acty_data5, aes(interval, steps, group = 1)) +
  geom_point() + geom_line() + facet_wrap(~weekday_type, ncol=1) +
  labs(x = "Interval", y = "Number of Steps", 
       title = "Average number of steps per interval across weekdays and weekends")
```
